---
title: "Untitled"
author: "Esen Duman"
date: "2025-01-09"
output: html_document
---

Install before rendering:

install.packages(c("httr", "zoo", "readr", "here", "renv", "data.table", "tidyverse", "ggplot2", "plotly", "dplyr", "stringr", "readxl","visNetwork", "caret" ))

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,      
  message = FALSE,      
  results = "hide",    
  fig.path = "figures/" 
)
library(data.table)
library(tidyverse)
library(plotly)
library(dplyr)
library(stringr)
library(httr)
library(here)
library(readr)
library(tidyr)
library(zoo)
library(ggplot2)
library(visNetwork)
library(caret)
```

# INTRODUCTION

In this analysis, we aim to explore various socioeconomic indicators such as Human Development Index (HDI), life expectancy, education levels, political regimes, and economic measures like Gross National Income (GNI) per capita. The objective is to clean, merge, and analyze datasets from multiple sources, primarily obtained from the World Bank, filling in missing values using regression-based imputation, and derive insights through visualizations and statistical models. Additionally, we utilize decision trees to identify key variables contributing to changes in GNI. 

# DATA COLLECTION

### 1. HDI

We downloaded the Human Development Index (HDI) dataset from the UNDP website and saved the file locally in a directory called data, before loading the dataset into R using `fread()` from the data.table package. We selected columns that start with hdi_ followed by a year, as our aim was to obtain each country's HDI by year, along with identifier columns and hdiscore column. Then, we transformed the dataset into a long format using the melt() function.

We had to change the format of the table, while keeping country and hdiscore columns, we created two new columns: 
Year is extracted from the column names and 
HDI is the HDI score for the corresponding year.

We renamed the country column to Country Name for consistency.
Lastly, we grouped the data by hdicode in 4 categories: "Low," "Medium," "High," "Very High". We will be using these categories in our further analysis, especially for visualization. 

A summary table showing the number of countries in each hdicode group is displayed in the output.

```{r}
url_hdi <- "https://hdr.undp.org/sites/default/files/2023-24_HDR/HDR23-24_Composite_indices_complete_time_series.csv"

file_path_hdi <- "data/hdi_time_series.csv"

dir.create("data", showWarnings = FALSE)

download.file(url_hdi, destfile = file_path_hdi, mode = "wb")

hdi <- fread(file_path_hdi, encoding = "UTF-8")

hdi_columns <- grep("^hdi_\\d+", colnames(hdi), value = TRUE)

hdi <- hdi[, c("country", "hdicode", hdi_columns), with = FALSE]

hdi_columns <- grep("^hdi_\\d+", colnames(hdi), value = TRUE)

hdi_long <- melt(
  hdi,
  id.vars = c("country", "hdicode"),    
  measure.vars = hdi_columns,                
  variable.name = "Year",                     
  value.name = "HDI")                          

hdi_long[, Year := as.numeric(gsub("hdi_", "", Year))]

colnames(hdi_long)[colnames(hdi_long) == "country"] <- "Country Name"

head(hdi_long)
```

### 2.Life Expectancy

The dataset is sourced from the World Bank API. First, we downloaded the data in a ZIP file format, extracted, and loaded it into R using the `fread()` function. Then, we cleaned the column names and assigned them properly by extracting the first row. We excluded irrelevant columns, and reshaped the dataset into a tidy format using `pivot_longer()`. With this transformation, we wanted to combine all year columns into a single Year column, while storing corresponding life expectancy values in a LifeExpectancy column. Additionally, we converted the Year column into integer form for proper numerical processing. The resulting dataset provides a structured format where each row represents a specific country, year, and its respective life expectancy value.

```{r}
url_LE <- "https://api.worldbank.org/v2/en/indicator/SP.DYN.LE00.IN?downloadformat=csv"

output_zip_LE <- here("data", "life_expectancy.zip")

dir.create("data", showWarnings = FALSE)

GET(url_LE, write_disk(output_zip_LE, overwrite = TRUE))

unzip(output_zip_LE, exdir = here("data"))

file_path_LE<- "data/API_SP.DYN.LE00.IN_DS2_en_csv_v2_87.csv"
life_expectancy_data_raw <- data.table::fread(file_path_LE, encoding = "UTF-8")

new_colnames <- as.character(unlist(life_expectancy_data_raw[1, ]))

setnames(life_expectancy_data_raw, new_colnames)

life_expectancy_data_raw <- life_expectancy_data_raw[-1, ]

setnames(life_expectancy_data_raw, gsub("\\.00000$", "", colnames(life_expectancy_data_raw)))

head(life_expectancy_data_raw)

life_expectancy <- life_expectancy_data_raw[,-c(2,3,4,68,69)]
head(life_expectancy)

tidy_data <- life_expectancy %>%
  pivot_longer(
    cols = starts_with("19") | starts_with("20"), 
    names_to = "Year",                           
    values_to = "LifeExpectancy"                 
  ) %>%
  mutate(Year = as.integer(Year))              

head(tidy_data)
```

### 3. Expected Years of Schooling

The dataset for expected years of schooling is sourced from the World Bank's gender data portal. The data is downloaded as a ZIP file and extracted into a local directory. We loaded the extracted CSV file into R and removed unnecessary columns to focus on the relevant data. There were a lot of missing values in the data. So, by using the zoo package, we filled the missing values with the previous non-missing value for consistency. Finally, we filtered the rows where the schooling data is categorized as "total", and renamed the "Value" column to Expected_Schooling for better clarity and consistency among our datasets.

```{r}
library(httr)  
library(zip)  

url_ee <- "https://extdataportal.worldbank.org/content/dam/sites/data/gender-data/data/data-gen/zip/indicator/expected-years-of-schooling.zip"
output_zip_ee <- "expected_years_of_schooling.zip"
output_dir_ee <- "expected_years_of_schooling"

download.file(url_ee, output_zip_ee, mode = "wb")

if (!dir.exists(output_dir_ee)) {
  dir.create(output_dir_ee)
}
unzip(output_zip_ee, exdir = output_dir_ee)

extracted_files <- list.files(output_dir_ee, full.names = TRUE)
print(extracted_files)

expected_years_of_schooling_raw<- read.csv("expected_years_of_schooling/Expected years of schooling.csv")
head(expected_years_of_schooling_raw)

expected_years_of_schooling <- expected_years_of_schooling_raw[,-c(1,2,4)]
head(expected_years_of_schooling)
```
```{r}

fill_na_with_previous <- function(x) {
  if (is.factor(x)) x <- as.character(x)
  return(na.locf(x, na.rm = FALSE, fromLast = FALSE))
}

library(zoo) 
expected_years_of_schooling <- expected_years_of_schooling %>%
  mutate(across(everything(), ~ na.locf(., na.rm = FALSE)))

expected_years_of_schooling_total<- expected_years_of_schooling %>% filter(Disaggregation == "total")
expected_years_of_schooling_total[["Year"]] <- as.numeric(as.character(expected_years_of_schooling_total[["Year"]]))
colnames(expected_years_of_schooling_total)[colnames(expected_years_of_schooling_total) == "Value"] <- "Expected_Schooling"
colnames(expected_years_of_schooling_total)[colnames(expected_years_of_schooling_total) == "Country.Name"] <- "Country Name"

head(expected_years_of_schooling_total)


```

### 4. GNI per Capita

The Gross National Income (GNI) per capita data is obtained through the World Bank API. We downloaded the data as a CSV file within a ZIP archive. After extraction, we loaded the data into R using fread(). Then, we cleaned and standardized column. We reshaped the dataset into a long format using the melt function, creating a structure where each row represents a country, a year, and its corresponding GNI per capita value to allow for easier integration with other datasets.

```{r}
url_gni <- "https://api.worldbank.org/v2/en/indicator/NY.GNP.PCAP.CD?downloadformat=csv"

output_zip_gni <- here("data3", "gni.zip")

dir.create("data3", showWarnings = FALSE)

GET(url_gni, write_disk(output_zip_gni, overwrite = TRUE))

unzip(output_zip_gni, exdir = here("data3"))

file_path_gni<- "data3/API_NY.GNP.PCAP.CD_DS2_en_csv_v2_4531.csv"
gni_data_raw <- data.table::fread(file_path_gni, encoding = "UTF-8")

new_colnames <- as.character(unlist(gni_data_raw[1, ]))

setnames(gni_data_raw, new_colnames)
gni_data_raw <- gni_data_raw[-1, ]

setnames(gni_data_raw, gsub("\\.00000$", "", colnames(gni_data_raw)))

head(gni_data_raw)

columns_to_keep <- c(
  "Country Name",
  grep("^197[0-9]|^198[0-9]|^199[0-9]|^200[0-9]|^201[0-9]|^202[0-9]$",
       colnames(gni_data_raw), value = TRUE)
)

gni_data_filtered <- gni_data_raw[, ..columns_to_keep]

gni_data_long <- melt(
  gni_data_filtered,
  id.vars = c("Country Name"),
  variable.name = "Year",   # Year column will retain exact year values (e.g., 1970, 1971, etc.)
  value.name = "GNI_Per_Capita"
)

gni_data_long[, Year := as.numeric(as.character(Year))]

head(gni_data_long)

data.table(gni_data_long)
```

### 5. Mean Years of Schooling

The mean years of schooling dataset is downloaded as a CSV file from Our World in Data. After loading the file into R, we again removed the unnecessary columns to retain only the country name, year, and schooling data.

```{r}
library(readxl)

url_ME <- "https://ourworldindata.org/grapher/mean-years-of-schooling-long-run.csv?v=1&csvType=full&useColumnShortNames=true"

file_path_ME <- "mean_years_of_schooling.csv"
download.file(url_ME, file_path_ME)
mean_schooling_raw <- read.csv(file_path_ME)
colnames(mean_schooling_raw) <- c("Country Name", "Code", "Year", "MeanYearsOfSchooling")

mean_schooling <- mean_schooling_raw[,-2]

head(mean_schooling)
```

### 6. Democracy

The political regime dataset, indicating the type of governance in countries over time, is acquired also from Our World in Data. The file is downloaded as a CSV and loaded into R. We removed unnecessary columns to only leave the country name, year, and political regime.
The classification distinguishes between closed autocracies (score 0), electoral autocracies (score 1), electoral democracies (score 2), and liberal democracies (score 3). In closed autocracies, citizens do not have the right to either choose the chief executive of the government or the legislature through multi-party elections. In electoral autocracies, citizens have the right to choose the chief executive and the legislature through multi-party elections; but they lack some freedoms, such as the freedoms of association or expression, that make the elections meaningful, free, and fair. In electoral democracies, citizens have the right to participate in meaningful, free and fair, and multi-party elections. In liberal demoracies, citizens have further individual and minority rights, are equal before the law, and the actions of the executive are constrained by the legislative and the courts.

```{r}
url_D <- "https://ourworldindata.org/grapher/political-regime.csv?v=1&csvType=full&useColumnShortNames=false"

file_path_D <- "democracy.csv"

download.file(url_D, file_path_D)

democracy_raw <- read.csv(file_path_D)
colnames(democracy_raw) <- c("Country Name", "Code", "Year", "Political.regime")

democracy <- democracy_raw[,-2]
head(democracy)
```

### 7. Gender Inequality

The gender inequality index dataset is sourced from the UN Human Development Report through Our World in Data. After downloading the CSV file, the data is loaded into R. Similar to other datasets, irrelevant columns are removed, leaving the country name, year, and gender inequality values.

```{r}
url_GI <- "https://ourworldindata.org/grapher/gender-inequality-index-from-the-human-development-report.csv?v=1&csvType=full&useColumnShortNames=false"

file_path_GI <- "democracy.csv"

download.file(url_GI, file_path_GI)

gender_inequality_raw <- read.csv(file_path_GI)
colnames(gender_inequality_raw) <- c("Country Name", "Code", "Year", "Gender.Inequality")

gender_inequality <- gender_inequality_raw[,-2]
head(gender_inequality)
```

### 8. Population

The population dataset, reflecting the population size of countries over time, is acquired from Our World in Data. The CSV file is downloaded, loaded into R, and cleaned to exclude unnecessary columns.

```{r}
url_pop <- "https://ourworldindata.org/grapher/population.csv?v=1&csvType=full&useColumnShortNames=true"

file_path_pop <- "population.csv"

download.file(url_pop, file_path_pop)

population_raw <- read.csv(file_path_pop)
colnames(population_raw) <- c("Country Name", "Code", "Year", "Population")

population <- population_raw[,-2] %>% filter(Year > 1960)
head(population)
```

### 9. Foreign Direct Investment

The foreign direct investment (FDI) dataset, showing net inflows as a percentage of GDP, is obtained from Our World in Data. The CSV file is downloaded and loaded into R. Again, we dropped irrelevant columns and kept only the essential country name, year, and FDI values. This dataset highlights the economic integration of countries through international investments.

The following variables are extracted in the same way as Foreign Direct Investment.

```{r}
url_fdi <- "https://ourworldindata.org/grapher/foreign-direct-investment-net-inflows-as-share-of-gdp.csv?v=1&csvType=full&useColumnShortNames=false"

file_path_fdi <- "FDI.csv"

download.file(url_fdi, file_path_fdi)

fdi_raw <- read.csv(file_path_fdi)
colnames(fdi_raw) <- c("Country Name", "Code", "Year", "FDI.inflow")

fdi <- fdi_raw[,-2]
head(fdi)
```

### 10. Tax Revenues 

```{r}
url_tax <- "https://ourworldindata.org/grapher/tax-revenues-as-a-share-of-gdp-unu-wider.csv?v=1&csvType=full&useColumnShortNames=false"

file_path_tax <- "tax_revenues.csv"

download.file(url_tax, file_path_tax)

tax_revenues_raw <- read.csv(file_path_tax)
colnames(tax_revenues_raw) <- c("Country Name", "Code", "Year", "Tax_Share")

tax_revenues <- tax_revenues_raw[,-2]
head(tax_revenues)
```

### 11. Electricity Generation
```{r}
url_EG <- "https://ourworldindata.org/grapher/per-capita-electricity-generation.csv?v=1&csvType=full&useColumnShortNames=true"

file_path_EG <- "electricity_generation.csv"

download.file(url_EG, file_path_EG)

electricity_generation_raw <- read.csv(file_path_EG)
colnames(electricity_generation_raw) <- c("Country Name", "Code", "Year", "Electricity_Generation")

electricity_generation <- electricity_generation_raw[,-2]
head(electricity_generation)
```

### 12. GDP per Capita
```{r}
url_GDP <- "https://ourworldindata.org/grapher/gdp-per-capita-worldbank.csv?v=1&csvType=full&useColumnShortNames=true"

file_path_GDP <- "electricity_generation.csv"

download.file(url_GDP, file_path_GDP)

GDPpercapita_raw <- read.csv(file_path_GDP)
colnames(GDPpercapita_raw) <- c("Country Name", "Code", "Year", "GDP_per_capita")

GDPpercapita <- GDPpercapita_raw[,-2]
head(GDPpercapita)
```

# Merging All Data

We prepared a comprehensive dataset (new_data) by merging all of our data tables. First, we used the `reduce` function to merge datasets based on common columns, Country Name and Year, ensuring a complete, unified dataset. The data is then filtered to include only the years between 1970 and 2023. 

Additionally, we cleaned the dataset by removing non-country entities such as regions (e.g., "Africa (UN)", "World", "Latin America & Caribbean") to ensure that only individual countries are analyzed. To address inconsistencies in country naming conventions across datasets, we used a mapping strategy with the `recode` function to standardize country names (e.g., Congo (Democratic Republic of the)" to "Congo"). Afterwards, we grouped the data by Country Name and Year and consolidated any duplicate rows. Our final dataset, new_data, is structured and cleaned, and is ready for further analysis.

```{r}
tables <- list(
  electricity_generation,
  tax_revenues,
  fdi,
  population,
  gender_inequality,
  democracy,
  mean_schooling,
  gni_data_long,
  expected_years_of_schooling_total,
  tidy_data,
  hdi_long,
  GDPpercapita
)

merged_data <- Reduce(function(x, y) merge(x, y, by = c("Country Name", "Year"), all = TRUE), tables)

merged_data <- subset(merged_data, Year >= 1970 & Year <= 2023)

str(merged_data)

merged_data <- merged_data[!merged_data$`Country Name` %in% c(
"Africa (UN)",                                         
"Africa Eastern and Southern" ,                        
"Africa Western and Central"   ,                       
"Americas (UN)"                 ,                      
"Arab States"                    ,                     
"Arab States (UNDP)"              ,                    
"Arab World"                       ,                   
"Asia"                              ,                  
"Asia (excl. China and India)"       ,                 
"Asia (UN)"                           ,                
"Caribbean small states"               ,               
"Central Europe and the Baltics"        ,              
 "Early-demographic dividend"            ,              
"East Asia & Pacific"                     ,            
 "East Asia & Pacific (excluding high income)"  ,       
  "East Asia & Pacific (IDA & IBRD countries)"   ,       
 "East Asia & Pacific (IDA & IBRD)"               ,     
 "East Asia and Pacific (WB)"                      ,    
 "East Asia and the Pacific"                        ,   
 "East Asia and the Pacific (UNDP)"                  ,  
 "East Germany"                                       , 
 "East Timor"                                          ,
 "Euro area"                                           ,
 "Europe"                                              ,
"Europe (excl. Russia)"                               ,
 "Europe (UN)"                                         ,
 "Europe & Central Asia"                               ,
 "Europe & Central Asia (excluding high income)"       ,
 "Europe & Central Asia (IDA & IBRD countries)"        ,
 "Europe & Central Asia (IDA & IBRD)"                  ,
 "Europe and Central Asia"                             ,
 "Europe and Central Asia (UNDP)"                      ,
 "Europe and Central Asia (WB)"                        ,
 "European Union"                                      ,
 "European Union (27)"                                 ,
 "Fragile and conflict affected situations"            ,
 "Heavily indebted poor countries (HIPC)"              ,
 "High human development"                              ,
 "High human development (UNDP)"                       ,
 "High income"                                         ,
 "High-income countries"                               ,
"IBRD only"                                           ,
 "IDA & IBRD total"                                    ,
 "IDA blend"                                           ,
 "IDA only"                                            ,
 "IDA total"                                           ,
"Late-demographic dividend"                           ,
 "Latin America & Caribbean"                           ,
 "Latin America & Caribbean (excluding high income)"   ,
 "Latin America & Caribbean (IDA & IBRD)"              ,
"Latin America & the Caribbean (IDA & IBRD countries)",
"Latin America and Caribbean (WB)"                    ,
 "Latin America and the Caribbean"                     ,
 "Latin America and the Caribbean (UN)"                ,
 "Latin America and the Caribbean (UNDP)"              ,
"Least developed countries: UN classification"        ,
"Low & middle income"                                 ,
"Low human development"                               ,
"Low human development (UNDP)"                        ,
"Low income"                                          ,
"Low-income countries"                                ,
"Lower middle income"                                 ,
"Lower-middle-income countries"                       ,
"Medium human development"                            ,
"Medium human development (UNDP)"                     ,
"Middle East & North Africa"                          ,
"Middle East & North Africa (excluding high income)"  ,
"Middle East & North Africa (IDA & IBRD countries)"   ,
"Middle East & North Africa (IDA & IBRD)"             ,
"Middle East and North Africa (WB)"                  , 
"Middle income"                                       ,
"Middle-income countries"                             ,
"North America"                                       ,
"North America (WB)"                                  ,
"Northern America (UN)"                               ,
"Not classified"                                      ,
"OECD members"                                        ,
"Other small states"                                  ,
"Pacific island small states"                         ,
"Post-demographic dividend"                           ,
"Pre-demographic dividend"                            ,
"Reunion"                                             ,
"Small states"                                        ,
"South America"                                       ,
"South America (excl. Brazil)"                        ,
"South Asia"                                          ,
"South Asia (IDA & IBRD)"                             ,
"South Asia (UNDP)"                                   ,
"South Asia (WB)"                                     ,
"Sub-Saharan Africa"                                  ,
"Sub-Saharan Africa (excluding high income)"          ,
"Sub-Saharan Africa (IDA & IBRD countries)"           ,
"Sub-Saharan Africa (IDA & IBRD)"                     ,
"Sub-Saharan Africa (UNDP)"                           ,
"Sub-Saharan Africa (WB)"                             ,
"Very high human development"                         ,
"Very high human development (UNDP)"                  ,
"West Bank and Gaza"                                  ,
"Western Sahara"                                      ,
"Oceania" ,                            
"Oceania (UN)"  , 
"World",
"Upper middle income",
"Upper-middle-income countries"
), ]


country_name_mapping <- c(
  "Congo (Democratic Republic of the)" = "Congo, Dem. Rep.",
  "Congo, Rep." = "Congo, Dem. Rep.",
   "Democratic Republic of Congo" =  "Congo, Rep." ,
   "Democratic Republic of Congo" = "Congo",
  "Congo, Dem. Rep."="Congo",
  "Bahamas, The"="Bahamas",
    "C\xf4te d'Ivoire"="Cote d'Ivoire",
  "Czechia"="Czechoslovakia",
  "Gambia, The"= "Gambia" ,
  "Bolivia (Plurinational State of)"= "Bolivia",
  "Ethiopia (former)"="Ethiopia",
  "Egypt, Arab Rep."="Egypt",
  "Hong Kong"="Hong Kong SAR, China",
  "Hong Kong SAR, China" ="Hong Kong, China (SAR)",
  "Eswatini (Kingdom of)"="Eswatini", 
   "Iran (Islamic Republic of)"="Iran",
 "Iran, Islamic Rep."="Iran",
  "Korea (Democratic People's Rep. of)" = "Korea (Republic of)",
  "Korea (Republic of)"="Korea, Dem. People's Rep.",
  "Korea, Dem. People's Rep."=    "Korea, Rep.",
   "Micronesia (country)"="Micronesia (Federated States of)" ,
   "Micronesia (Federated States of)"="Micronesia, Fed. Sts.",
  "Micronesia, Fed. Sts."="Micronesia",
"Moldova (Republic of)"="Moldova" ,
"Kyrgyz Republic"= "Kyrgyzstan"    ,
"Lao PDR"="Lao People's Democratic Republic"   ,
 "Macao"="Macao SAR, China",
"Palestine"="Palestine, State of",
   "Palestine, State of"="Palestine/Gaza",
   "Palestine/Gaza" = "Palestine/West Bank",
"Palestine/West Bank"  = "Palestine/Gaza" ,
"Slovak Republic"="Slovakia"                    ,
"Yemen" ="Yemen Arab Republic",
   "Yemen Arab Republic"="Yemen People's Republic",
   "Yemen People's Republic"="Yemen, Rep.",
"Russia"="Russian Federation",
"Serbia"="Serbia and Montenegro"          ,
"Syrian Arab Republic" ="Syria"         ,
  "Viet Nam"= "Vietnam",     
  "Democratic Republic of Vietnam" = "Republic of Vietnam",
 "Tanzania (United Republic of)"="Tanzania" ,
"T\xfcrkiye" = "Turkey",
"Turkey"="Turkiye",
 "Venezuela (Bolivarian Republic of)" = "Venezuela",
"Venezuela, RB" = "Venezuela"

)
new_data <- merged_data %>%
  mutate(`Country Name` = recode(`Country Name`, !!!country_name_mapping))

country_name_mapping2 <- c(
  "Palestine/West Bank"  = "Palestine/Gaza",
  "Congo, Dem. Rep."    = "Congo",
  "Turkey" = "Turkiye",
   "Yemen, Rep." = "Yemen People's Republic",
  "Micronesia, Fed. Sts."    =  "Micronesia" ,
   "Korea, Rep."     = "Korea, Dem. People's Rep."  
)
  
new_data <- new_data %>%
  mutate(`Country Name` = recode(`Country Name`, !!!country_name_mapping2))


str(new_data)

new_data <- new_data %>%
  group_by(`Country Name`, Year) %>%
  summarize(across(everything(), ~ mean(., na.rm = TRUE)), .groups = "drop")

str(new_data)


```

The countries are grouped in the hdicode column by 4 categories: "Low," "Medium," "High," "Very High". Utilizing this classification, we selected 5 countries from each category and created a subset of the new_data dataset to use for our further analysis. We will mainly be using this dataset for our plots to observe the differences in many variables for each category.

```{r}
hdicode_group <- hdi_long[, .(
  Number_of_Countries = uniqueN(`Country Name`) 
), by = hdicode]

print(hdicode_group)

set.seed(213276)
low <- hdi_long[hdicode == "Low", sample(unique(`Country Name`), 5)]

set.seed(213276)
medium <- hdi_long[hdicode == "Medium", sample(unique(`Country Name`), 5)]

set.seed(213276)
high <- hdi_long[hdicode == "High", sample(unique(`Country Name`), 5)]

set.seed(213276)
very_high <- hdi_long[hdicode == "Very High", sample(unique(`Country Name`), 5)]

selected_countries <- new_data %>%
  filter(`Country Name` %in% c(low, medium, high, very_high))

selected_countries <- selected_countries %>%
  mutate(
    HDI_Group = case_when(
      `Country Name` %in% low ~ "Low",
      `Country Name` %in% medium ~ "Medium",
      `Country Name` %in% high ~ "High",
      `Country Name` %in% very_high ~ "Very High"
    )
  )
```

# PLOTS

```{r}
ggplot(new_data, aes(x = Year, y = LifeExpectancy, color = LifeExpectancy, group = `Country Name`)) +
  geom_line(alpha = 0.7) +
  scale_color_viridis_c() +
  theme_minimal()
```

```{r}
library(ggplot2)
library(ggrepel)

label_data <- selected_countries %>%
  group_by(`Country Name`, HDI_Group) %>%
  filter(Year == 2022)

ggplot(selected_countries, aes(x = Year, y = LifeExpectancy, color = `Country Name`, group = `Country Name`)) +
  geom_line(size = 0.5) +
  geom_text_repel(
    data = label_data,
    aes(label = `Country Name`),
    show.legend = FALSE,
    size = 2
  ) +
  facet_wrap(~ HDI_Group, ncol = 2, scales = "fixed") +  
  labs(
    title = "Life Expectancy Over Time by HDI Group",
    subtitle = "Selected Countries",
    x = "Year",
    y = "Life Expectancy"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "gray30"),
    legend.position = "none",  # Remove legend
    panel.grid.minor = element_blank()
  )

```


```{r}
new_data <- new_data %>% arrange(`Country Name`, Year)

create_interactive_plot <- function(data, group, title) {
  filtered_data <- data %>% filter(`Country Name` %in% group)

  plot <- ggplot(filtered_data, aes(
    x = Year, y = LifeExpectancy,
    color = `Country Name`,  
    group = `Country Name`, 
    text = paste("Country: ", `Country Name`, "<br>",
                 "Year: ", Year, "<br>",
                 "Life Expectancy: ", round(LifeExpectancy, 2))
  )) +
    geom_line(size = 0.5, alpha = 0.8) +  # Ensure proper line plotting
    theme_minimal() +
    scale_color_viridis_d() +  # Use discrete viridis colors for countries
    labs(
      title = title,
      x = "Year",
      y = "Life Expectancy (Years)"
    )

  ggplotly(plot, tooltip = "text")
}

interactive_low <- create_interactive_plot(
  new_data, low, "Life Expectancy Over Time in Countries with Low HDI Scores"
)
interactive_low

interactive_high <- create_interactive_plot(
  new_data, high, "Life Expectancy Over Time in Countries with High HDI Scores"
)
interactive_high

interactive_medium <- create_interactive_plot(
  new_data, medium, "Life Expectancy Over Time in Countries with Medium HDI Scores"
)
interactive_medium

interactive_very_high<- create_interactive_plot(
  new_data, very_high, "Life Expectancy Over Time in Countries with Very High HDI Scores"
)
interactive_very_high
```

```{r}
ggplot(new_data, aes(x = Year, y = Expected_Schooling, color = Expected_Schooling, group = `Country Name`)) +
  geom_line(alpha = 0.7) +
  scale_color_viridis_c() +
  theme_minimal()
```

```{r}
ggplot(selected_countries, aes(x = Year, y = Expected_Schooling, color = `Country Name`, group = `Country Name`)) +
  geom_line(size = 0.5) +
  geom_text_repel(data = label_data, aes(label = `Country Name`), hjust = -0.2, show.legend = TRUE) +
  facet_wrap(~ HDI_Group, ncol = 2, scales = "fixed") +  # 2x2 layout
  labs(
    title = "Expected Years of Schooling Over Time by HDI Group",
    subtitle = "Selected Countries",
    x = "Year",
    y = "Expected Years of Schooling",
    color = "Country"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "gray30"),
    legend.position = "none",  # Remove legend as labels are on the plot
    panel.grid.minor = element_blank()
  )
```

```{r}
library(ggplot2)
library(RColorBrewer)
library(dplyr)
library(ggrepel)

colors_very_high <- brewer.pal(5, "Reds")      
colors_high <- brewer.pal(5, "Blues")          
colors_medium <- brewer.pal(5, "Greens")       
colors_low <- brewer.pal(5, "Purples")         

country_colors <- c(
  setNames(colors_low, low),                   
  setNames(colors_medium, medium),             
  setNames(colors_high, high),                 
  setNames(colors_very_high, very_high)        
)

gni_data_long <- gni_data_long %>%
  mutate(
    Group = case_when(
      `Country Name` %in% low ~ "Low",
      `Country Name` %in% medium ~ "Medium",
      `Country Name` %in% high ~ "High",
      `Country Name` %in% very_high ~ "Very High",
      TRUE ~ NA_character_  # Explicitly set NA for unmatched countries
    )
  ) %>%
  filter(!is.na(Group))  # Remove rows with NA in Group

label_data2 <- gni_data_long %>%
  group_by(`Country Name`, Group) %>%
  filter(Year == max(Year))

ggplot(selected_countries, aes(x = Year, y = GNI_Per_Capita, color = `Country Name`, group = `Country Name`)) +
  geom_line(size = 0.8) +
  geom_text_repel(data = label_data2, aes(label = `Country Name`), hjust = -0.2, show.legend = FALSE) +
  facet_wrap(~ Group, scales = "free_y") +  # Separate facets for each group
  scale_color_manual(values = country_colors) +  # Apply custom colors
  theme_minimal() +
  labs(
    title = "GNI Per Capita Over Time by Group",
    x = "Year",
    y = "GNI Per Capita (USD)"
  ) +
  theme(
    legend.position = "none",  # Remove legend
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "gray30")
  )

```

```{r}
ggplot(mean_schooling, aes(x = Year, y = MeanYearsOfSchooling, color = MeanYearsOfSchooling, group = `Country Name`)) +
  geom_line(alpha = 0.7) +
  scale_color_viridis_c() +
  theme_minimal()
```


```{r}

str(democracy)

aggregated_data <- democracy %>%
  group_by(Year, Political.regime) %>%
  summarize(Count = n(), .groups = "drop")

ggplot(aggregated_data, aes(x = Year, y = Count, color = factor(Political.regime), group = Political.regime)) +
  geom_line(size = 0.5) +
  geom_point(size = 0.2) +  
  scale_color_viridis_d() +
  theme_minimal(base_size = 14) +
  labs(
    title = "Number of Countries by Political Regime Over Time",
    x = "Year",
    y = "Number of Countries",
    color = "Political Regime"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for readability
    legend.position = "right"  # Move the legend to the side
  )

```

```{r}

ggplot(selected_countries, aes(x = Year, y = Political.regime , color = `Country Name`, group = `Country Name`)) +
  geom_line(size = 0.5) +
  geom_text_repel(data = label_data, aes(label = `Country Name`), hjust = -0.2, show.legend = TRUE) +
  facet_wrap(~ HDI_Group, ncol = 2, scales = "fixed") +  # 2x2 layout
  labs(
    title = "Political Regimes Over Time by HDI Group",
    subtitle = "Selected Countries",
    x = "Year",
    y = "Political Regimes",
    color = "Country"
  ) +
  scale_y_continuous(
    breaks = 0:3,  # Add breaks for 0, 1, 2, 3
    labels = c(
      "0 = Closed Autocracy",
      "1 = Electoral Autocracy",
      "2 = Electoral Democracy",
      "3 = Liberal Democracy"
    )  # Add labels to explain values
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "gray30"),
    legend.position = "none",  # Remove legend for color (labels are on the plot)
    panel.grid.minor = element_blank()
  )
```

```{r}
ggplot(gender_inequality, aes(x = Year, y = Gender.Inequality, color = Gender.Inequality, group = `Country Name`)) +
  geom_line(alpha = 0.7) +
  scale_color_viridis_c() +
  theme_minimal()

```

```{r}
selected_countries_1990 <- selected_countries %>%
  filter(Year >= 1990)

ggplot(selected_countries_1990, aes(x = Year, y = Gender.Inequality, color = `Country Name`, group = `Country Name`)) +
  geom_line(size = 0.5) +
  geom_text_repel(data = label_data, aes(label = `Country Name`), hjust = -0.2, show.legend = TRUE) +
  facet_wrap(~ HDI_Group, ncol = 2, scales = "fixed") +  # 2x2 layout
  labs(
    title = "Gender Inequality Index Over Time by HDI Group",
    subtitle = "Selected Countries",
    x = "Year",
    y = "Gender Inequality Index",
    color = "Country"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "gray30"),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

```{r}
ggplot(population, aes(x = Year, y = Population, color = Population, group = `Country Name`)) +
  geom_line(alpha = 0.7) +
  scale_color_viridis_c() +
  theme_minimal()
```
```{r}
ggplot(selected_countries, aes(x = Year, y = Population, color = `Country Name`, group = `Country Name`)) +
  geom_line(size = 0.5) +
  geom_text_repel(data = label_data, aes(label = `Country Name`), hjust = -0.2, show.legend = TRUE) +
  facet_wrap(~ HDI_Group, ncol = 2, scales = "fixed") +  # 2x2 layout
  labs(
    title = "Population Over Time by HDI Group",
    subtitle = "Selected Countries",
    x = "Year",
    y = "Population",
    color = "Country"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "gray30"),
    legend.position = "none", 
    panel.grid.minor = element_blank()
  )
```


```{r}
ggplot(fdi, aes(x = Year, y = FDI.inflow, color = FDI.inflow, group = `Country Name`)) +
  geom_line(alpha = 0.7) +
  scale_color_viridis_c() +
  theme_minimal()
```
```{r}
ggplot(selected_countries, aes(x = Year, y = FDI.inflow, color = `Country Name`, group = `Country Name`)) +
  geom_line(size = 0.5) +
  geom_text_repel(data = label_data, aes(label = `Country Name`), hjust = -0.2, show.legend = TRUE) +
  facet_wrap(~ HDI_Group, ncol = 2, scales = "fixed") +  # 2x2 layout
  labs(
    title = "FDI Over Time by HDI Group",
    subtitle = "Selected Countries",
    x = "Year",
    y = "Foreign Direct Investment Net Inflows",
    color = "Country"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "gray30"),
    legend.position = "none", 
    panel.grid.minor = element_blank()
  )
```


```{r}
ggplot(new_data, aes(x = Year, y = GDP_per_capita, color = GDP_per_capita, group = `Country Name`)) +
  geom_line(alpha = 0.7) +
  scale_color_viridis_c() +
  theme_minimal()
```

```{r}
ggplot(selected_countries_1990, aes(x = Year, y = GDP_per_capita, color = `Country Name`, group = `Country Name`)) +
  geom_line(size = 0.5) +
  geom_text_repel(data = label_data, aes(label = `Country Name`), hjust = -0.2, show.legend = TRUE) +
  facet_wrap(~ HDI_Group, ncol = 2, scales = "fixed") +  # 2x2 layout
  labs(
    title = "GDP per Capita Over Time by HDI Group",
    subtitle = "Selected Countries",
    x = "Year",
    y = "GDP per Capita",
    color = "Country"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "gray30"),
    legend.position = "none", 
    panel.grid.minor = element_blank()
  )
```


```{r}
ggplot(tax_revenues, aes(x = Year, y = Tax_Share, color = Tax_Share, group = `Country Name`)) +
  geom_line(alpha = 0.7) +
  scale_color_viridis_c() +
  theme_minimal()
```
```{r}

selected_countries_1980 <- selected_countries %>%
  filter(Year >= 1980)

ggplot(selected_countries_1980, aes(x = Year, y = Tax_Share, color = `Country Name`, group = `Country Name`)) +
  geom_line(size = 0.5) +
  geom_text_repel(data = label_data, aes(label = `Country Name`), hjust = -0.2, show.legend = TRUE) +
  facet_wrap(~ HDI_Group, ncol = 2, scales = "fixed") +  # 2x2 layout
  labs(
    title = "Tax Revenue as a Share of GDP Over Time by HDI Group",
    subtitle = "Selected Countries",
    x = "Year",
    y = "Tax Share",
    color = "Country"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "gray30"),
    legend.position = "none", 
    panel.grid.minor = element_blank()
  )
```


```{r}
ggplot(electricity_generation, aes(x = Year, y = Electricity_Generation, color = Electricity_Generation, group = `Country Name`)) +
  geom_line(alpha = 0.7) +
  scale_color_viridis_c() +
  theme_minimal()
```

```{r}

ggplot(selected_countries_1990, aes(x = Year, y = Electricity_Generation, color = `Country Name`, group = `Country Name`)) +
  geom_line(size = 0.5) +
  geom_text_repel(data = label_data, aes(label = `Country Name`), hjust = -0.2, show.legend = TRUE) +
  facet_wrap(~ HDI_Group, ncol = 2, scales = "fixed") +  # 2x2 layout
  labs(
    title = "Electricity Generation Over Time by HDI Group",
    subtitle = "Selected Countries",
    x = "Year",
    y = "Electricity Generation",
    color = "Country"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "gray30"),
    legend.position = "none", 
    panel.grid.minor = element_blank()
  )
```

# Predicting Missing Values with a Linear Regression Model

There were a lot of missing values in our dataset. So, we wanted to predict some of these values by a regression model.

First, we subsetted our new_data dataset to create model_data, which only includes the necessary columns for our regression analysis. The selected columns are Country Name, Year, LifeExpectancy, GNI_Per_Capita, MeanYearsOfSchooling, Expected_Schooling, and HDI. 

We then used these columns to fit a linear regression model (model) where HDI is the dependent variable, and the rest are independent variables. The summary(model) function provides the regression results, including coefficients, residuals, and significance levels. 

From the model summary we can see that there are indeed a lot of missing values, 13314 observations were deleted due to missingness.

We identified the rows with exactly one missing value in the selected columns (rows_with_one_missing) for further analysis. These rows will be targeted for imputation using the coefficients from the regression model.

```{r}

necessary_columns <- c("Country Name", "Year", "LifeExpectancy", "GNI_Per_Capita", 
                       "MeanYearsOfSchooling", "Expected_Schooling", "HDI")

model_data <- new_data %>%
  select(all_of(necessary_columns))

model <- lm(HDI  ~ LifeExpectancy + GNI_Per_Capita + MeanYearsOfSchooling + Expected_Schooling, data=model_data )

summary(model)

columns_to_check <- c("LifeExpectancy", "GNI_Per_Capita", "MeanYearsOfSchooling", "Expected_Schooling", "HDI")

rows_with_one_missing <- model_data %>%
  rowwise() %>%
  mutate(missing_count = sum(is.na(c_across(all_of(columns_to_check))))) %>%
  filter(missing_count == 1) %>%
  ungroup()

head(rows_with_one_missing)
nrow(rows_with_one_missing)
coefficients <- coef(model)
```

We created a custom function, predict_or_calculate_hdi, to predict missing values in the dataset. The function uses the regression coefficients obtained from the model to calculate the predicted value based on the available independent variables. 

The function is applied row-by-row to the dataset rows_with_one_missing. For rows with missing values in one column, the corresponding value is predicted and replaced. 

The result is a dataset completed_rows, where missing values have been imputed.

```{r}
predict_or_calculate_hdi <- function(row, coefficients) {
  intercept <- coefficients["(Intercept)"]
  
  predicted_value <- intercept
  
  if (!is.na(row$LifeExpectancy)) {
    predicted_value <- predicted_value + coefficients["LifeExpectancy"] * row$LifeExpectancy
  }
  if (!is.na(row$GNI_Per_Capita)) {
    predicted_value <- predicted_value + coefficients["GNI_Per_Capita"] * row$GNI_Per_Capita
  }
  if (!is.na(row$MeanYearsOfSchooling)) {
    predicted_value <- predicted_value + coefficients["MeanYearsOfSchooling"] * row$MeanYearsOfSchooling
  }
  if (!is.na(row$Expected_Schooling)) {
    predicted_value <- predicted_value + coefficients["Expected_Schooling"] * row$Expected_Schooling
  }
  
  return(predicted_value)
}

completed_rows <- rows_with_one_missing %>%
  rowwise() %>%
  mutate(
    LifeExpectancy = ifelse(
      is.na(LifeExpectancy),
      predict_or_calculate_hdi(cur_data(), coefficients),
      LifeExpectancy
    ),
    GNI_Per_Capita = ifelse(
      is.na(GNI_Per_Capita),
      predict_or_calculate_hdi(cur_data(), coefficients),
      GNI_Per_Capita
    ),
    MeanYearsOfSchooling = ifelse(
      is.na(MeanYearsOfSchooling),
      predict_or_calculate_hdi(cur_data(), coefficients),
      MeanYearsOfSchooling
    ),
    Expected_Schooling = ifelse(
      is.na(Expected_Schooling),
      predict_or_calculate_hdi(cur_data(), coefficients),
      Expected_Schooling
    ),
    HDI = ifelse(
      is.na(HDI),
      predict_or_calculate_hdi(cur_data(), coefficients),
      HDI
    )
  ) %>%
  ungroup()

head(completed_rows)
```

We merged the predicted values from completed_rows back into model_data to create a new dataset, model_data_with_predictions. Here, we used the `left_join` function to combine the original and predicted datasets based on Country Name and Year. 
The result is a dataset with imputed values integrated into the original dataset. 

We conducted a new regression model, model2, fitted on this updated dataset to evaluate the impact of the imputed values on the regression analysis. We saw that , while conducting this analysis, 10310 observations were deleted, meaning that around 3000 observations were imputed by our function. 

```{r}
library(dplyr)

model_data_with_predictions <- model_data %>%
  left_join(
    completed_rows %>%
      select(`Country Name`, Year, LifeExpectancy, GNI_Per_Capita, MeanYearsOfSchooling, Expected_Schooling, HDI),
    by = c("Country Name", "Year"),
    suffix = c("_original", "_predicted")
  ) %>%
  mutate(
    LifeExpectancy = ifelse(is.na(LifeExpectancy_original), LifeExpectancy_predicted, LifeExpectancy_original),
    GNI_Per_Capita = ifelse(is.na(GNI_Per_Capita_original), GNI_Per_Capita_predicted, GNI_Per_Capita_original),
    MeanYearsOfSchooling = ifelse(is.na(MeanYearsOfSchooling_original), MeanYearsOfSchooling_predicted, MeanYearsOfSchooling_original),
    Expected_Schooling = ifelse(is.na(Expected_Schooling_original), Expected_Schooling_predicted, Expected_Schooling_original),
    HDI = ifelse(is.na(HDI_original), HDI_predicted, HDI_original)
  ) %>%
  select(-ends_with("_original"), -ends_with("_predicted"))


str(model_data_with_predictions)

model2 <- lm(HDI  ~ LifeExpectancy + GNI_Per_Capita + MeanYearsOfSchooling + Expected_Schooling, data=model_data_with_predictions )
summary(model2)
```

Lastly, we wanted to visualize the number of imputed observations for each column. We created a bar plot, showing the number of cells filled for each column. 

```{r}
library(dplyr)
library(ggplot2)

filled_counts <- rows_with_one_missing %>%
  summarise(
    LifeExpectancy = sum(is.na(LifeExpectancy) & !is.na(completed_rows$LifeExpectancy)),
    GNI_Per_Capita = sum(is.na(GNI_Per_Capita) & !is.na(completed_rows$GNI_Per_Capita)),
    MeanYearsOfSchooling = sum(is.na(MeanYearsOfSchooling) & !is.na(completed_rows$MeanYearsOfSchooling)),
    Expected_Schooling = sum(is.na(Expected_Schooling) & !is.na(completed_rows$Expected_Schooling)),
    HDI = sum(is.na(HDI) & !is.na(completed_rows$HDI))
  ) %>%
  pivot_longer(cols = everything(), names_to = "Column", values_to = "FilledCount")

ggplot(filled_counts, aes(x = Column, y = FilledCount, fill = Column)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Number of Cells Filled for Each Column",
    x = "Column",
    y = "Number of Cells Filled",
    fill = "Column"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```

# Decision Trees

### Separating GDP into 5 Buckets

At first we tried to predict GNI, however GNI data was very limited. Therefore, we chaged our approach and decided to use GDP per capita as a mesure of wealth of the country. 

We created 4 buckets of GDP per capita. We used percentiles to separate GDP per capita into equel size buckets.We categorized GDP per capita in countries in different years as low, lower middle, upper middle and high.

```{r}
percentiles <- quantile(new_data$GDP_per_capita, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)
new_data$GDP_Bucket <- cut(
  new_data$GDP_per_capita,
  breaks = percentiles,
  labels = c("Low", "Lower-Middle", "Upper-Middle-Low", "High"),
  include.lowest = TRUE
)

table(new_data$GDP_Bucket)
```

#### Baseline Guessing

To be able to compare our decision tree models' performances, we printed the random guessing model's accuracy. 

```{r} 
baseline_random <- 1 / 4
print(paste("Baseline Accuracy (Random Guessing):", baseline_random))
```

## Decision Tree with only HDI Variables

We wanted to see if the factors in HDI like life expectancy and education has a role on GNI. Therefore, we regressed GDP classes on mean years of schooling, expected years of schooling and life expectancy. 

```{r}
set.seed(375)  
split <- sample(1:nrow(new_data), size = 0.7 * nrow(new_data))
train_data <- new_data[split, ]
test_data <- new_data[-split, ]
```

```{r}
library(rpart.plot)
library(rpart)
tree_model_HDI <- rpart(
  GDP_Bucket ~ MeanYearsOfSchooling + Expected_Schooling + LifeExpectancy,
  data = train_data,
  method = "class", 
  #control = rpart.control(minsplit = 5)
)

rpart.plot(tree_model_HDI)

visNetwork::visTree(tree_model_HDI)

```

The most prominant factor determining the GDP per capita is life expectancy. However, this only shows us a strong correlation between life expectancy and GDP per capita. We can not determine the direction of the relationship from this decision tree. 

```{r}
predictions_HDI <- predict(tree_model_HDI, test_data, type = "class")
confusion_matrix_HDI <- caret::confusionMatrix(table(fitted = predictions_HDI, actual = test_data$GDP_Bucket))

cat(capture.output(confusion_matrix_HDI), sep = "\n")
print(confusion_matrix_HDI)
```

Accuracy is 0.63 which a lot higher than baseline accuracy. Therefore, our model improves the predictions. However, its success is limited. 
We are going to incorparate more variables into our model to improve its prediction power.

## Continuous GDP Prediction

Without separating GDP into buckets, we used a continuous decision tree model to predict the exact value of GDP per capita.
We wanted to see if there are any differences on most effective factors between continuous and discrete GDP per capita values. 

### Decision Tree with All variables minus Electricty Generation

```{r}
library(rpart.plot)
library(rpart)
tree_model_all <- rpart(
  GDP_per_capita ~  Tax_Share + Population +Gender.Inequality + Political.regime + MeanYearsOfSchooling + Expected_Schooling + LifeExpectancy + FDI.inflow,
  data = train_data,
  method = "anova", 
  #control = rpart.control(minsplit = 5)
)

rpart.plot(tree_model_all)


visNetwork::visTree(tree_model_all)
```

#### Decision Tree with All variables after 2000

Electricity generation data before 2000 was limited for most of the countries. Therefore, we wanted to train another decision tree model with the data after 2000 to see if electricity generation is an important factor of GDP per capita.

```{r}
library(dplyr)
library(tidyr)

data_after_2000 <- new_data %>%
  filter(Year >= 2000)

head(data_after_2000)
```

We separated the data after 2000 into training and test sets.

```{r}
set.seed(389)  
split_2000 <- sample(1:nrow(data_after_2000), size = 0.7 * nrow(data_after_2000))
train_data_after2000 <- data_after_2000[split_2000, ]
test_data_after2000 <- data_after_2000[-split_2000, ]
```

```{r}
tree_model_2000 <- rpart(
  GNI_Per_Capita ~  Tax_Share + Population + Electricity_Generation + Gender.Inequality + Political.regime + MeanYearsOfSchooling +
    Expected_Schooling + LifeExpectancy + FDI.inflow,
  data = train_data_after2000,
  method = "anova", 
  #control = rpart.control(minsplit = 5)
)
rpart.plot(tree_model_2000)

visNetwork::visTree(tree_model_2000)
```

Using the data after 2000 showed us that the most important factor to determine GDP per capita is population. After that electricity generation is the factor that decreases the entropy the most. Higher electricity generation indicated a higher GDP per capita. This is perfectly logical since most of the advance economies use high amounts of electricity in production and technology services. 

## Decision Tree with 4 GDP Buckets

Now instead of using a continuous model, we are going to use a discrete classification model with 4 groups of the target variable. 

### With All the Data

```{r}
library(rpart)
library(rpart.plot)

tree_model_buckets <- rpart(
  GDP_Bucket ~ Electricity_Generation + Tax_Share + Population + Gender.Inequality + Political.regime + MeanYearsOfSchooling + Expected_Schooling + LifeExpectancy ,
  data = train_data,
  method = "class", 
  control = rpart.control(minsplit = 20, cp = 0.01) )

rpart.plot(tree_model_buckets, type = 3, extra = 104, box.palette = "Blues", shadow.col = "gray")

```

```{r}
visNetwork::visTree(tree_model_buckets)
```

In the discrete model, most important factor is life expectancy. We now that most developed nations have higher life expectancies therefore higher life expectancy points to a higher GDP. Again we see the importance of electricity generation at it is the second and third node.
The significant difference between discrete and continuous model is that in the discrete one, population is not that important to determine GDP as it is the factor in the last node.

```{r}
predictions_buckets_all <- predict(tree_model_buckets, test_data, type = "class")
confusion_matrix_buckets_all <- caret::confusionMatrix(table(fitted = predictions_buckets_all, actual = test_data$GDP_Bucket))
cat(capture.output(confusion_matrix_buckets_all), sep = "\n")
print(confusion_matrix_buckets_all)
```

We see that accuracy increased to 73%. It increased 10 percentage point compared to using only HDI factors. Incorparating more factors into the model allows us to make better GDP per capita predictions.

### After 2000 with buckets of GDP 

We are going to do a decision tree with all data after 2000 to predict the group of GDP per capita.

```{r}
tree_model_buckets_2000 <- rpart(
  GDP_Bucket ~ Electricity_Generation + Tax_Share + Population + Gender.Inequality + Political.regime + MeanYearsOfSchooling + Expected_Schooling + LifeExpectancy ,
  data = train_data_after2000,
  method = "class", 
  control = rpart.control(minsplit = 20, cp = 0.01)  # 
)

visNetwork::visTree(tree_model_buckets_2000)
```

Again, countries producing high electricity are classified as high GDP per capita countries. And high life expectancy points to a high GDP. 
The effect of political regime is an interesting one. As Daron Acemolu mentioned in his books, democracy is an important factor in development and GDP. He said that countries with storng democracies are generally richer. However in our model the effet of political regime is the opposite one. Countries with 0 in their political regimes are autocracies. And in this model, 0 in political regime points to a high GDP.  

```{r}
predictions_buckets_2000 <- predict(tree_model_buckets, test_data_after2000, type = "class")
confusion_matrix_buckets_2000 <- caret::confusionMatrix(table(fitted = predictions_buckets_2000, actual = test_data_after2000$GDP_Bucket))

cat(capture.output(confusion_matrix_buckets_2000), sep = "\n")
print(confusion_matrix_buckets_2000)
```

The accuracy of the model that is trained with the data after 2000 is higher with 75% score. This may be due to the fact that the data quality is higher and there are much less unavaliable data points after 2000. Therefore, the model can make better predictions and has a lot more data to learn from.