---
title: "Untitled"
author: "Esen Duman"
date: "2025-01-09"
output: html_document
---

Install before rendering:

install.packages(c("httr", "zoo", "readr", "here", "renv", "data.table", "tidyverse", "ggplot2", "plotly", "dplyr", "stringr", "readxl","visNetwork", "caret" ))

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,      # Suppress warnings
  message = FALSE,      # Suppress messages
  results = "hide",     # Suppress textual results in code chunks (optional)
  fig.path = "figures/" # Save figures in the "figures" folder
)
install.packages("here")
library(data.table)
library(tidyverse)
library(plotly)
library(dplyr)
library(stringr)
library(httr)
library(here)
library(readr)
library(tidyr)
library(zoo)
library(ggplot2)
library(visNetwork)
library(caret)
```

#INTRODUCTION

In this analysis, we aim to explore various socioeconomic indicators such as Human Development Index (HDI), life expectancy, education levels, political regimes, and economic measures like Gross National Income (GNI) per capita. The objective is to clean, merge, and analyze datasets from multiple sources, primarily obtained from the World Bank, filling in missing values using regression-based imputation, and derive insights through visualizations and statistical models. Additionally, we utilize decision trees to identify key variables contributing to changes in GNI. 

#DATA COLLECTION

###1. HDI

We downloaded the Human Development Index (HDI) dataset from the UNDP website and saved the file locally in a directory called data, before loading the dataset into R using `fread()` from the data.table package. We selected columns that start with hdi_ followed by a year, as our aim was to obtain each country's HDI by year, along with identifier columns and hdiscore column. Then, we transformed the dataset into a long format using the melt() function.

We had to change the format of the table, while keeping country and hdiscore columns, we created two new columns: 
Year is extracted from the column names and 
HDI is the HDI score for the corresponding year.

We renamed the country column to Country Name for consistency.
Lastly, we grouped the data by hdicode in 4 categories: "Low," "Medium," "High," "Very High". We will be using these categories in our further analysis, especially for visualization. 

A summary table showing the number of countries in each hdicode group is displayed in the output.

```{r}
# Define the URL
url_hdi <- "https://hdr.undp.org/sites/default/files/2023-24_HDR/HDR23-24_Composite_indices_complete_time_series.csv"

file_path_hdi <- "data/hdi_time_series.csv"

dir.create("data", showWarnings = FALSE)

download.file(url_hdi, destfile = file_path_hdi, mode = "wb")

hdi <- fread(file_path_hdi, encoding = "UTF-8")

hdi_columns <- grep("^hdi_\\d+", colnames(hdi), value = TRUE)

hdi <- hdi[, c("country", "hdicode", hdi_columns), with = FALSE]

hdi_columns <- grep("^hdi_\\d+", colnames(hdi), value = TRUE)

hdi_long <- melt(
  hdi,
  id.vars = c("country", "hdicode"),    
  measure.vars = hdi_columns,                
  variable.name = "Year",                     
  value.name = "HDI")                          

hdi_long[, Year := as.numeric(gsub("hdi_", "", Year))]

colnames(hdi_long)[colnames(hdi_long) == "country"] <- "Country Name"

head(hdi_long)
```

###2.Life Expectancy

The dataset is sourced from the World Bank API. First, we downloaded the data in a ZIP file format, extracted, and loaded it into R using the `fread()` function. Then, we cleaned the column names and assigned them properly by extracting the first row. We excluded irrelevant columns, and reshaped the dataset into a tidy format using `pivot_longer()`. With this transformation, we wanted to combine all year columns into a single Year column, while storing corresponding life expectancy values in a LifeExpectancy column. Additionally, we converted the Year column into integer form for proper numerical processing. The resulting dataset provides a structured format where each row represents a specific country, year, and its respective life expectancy value.

```{r}
url_LE <- "https://api.worldbank.org/v2/en/indicator/SP.DYN.LE00.IN?downloadformat=csv"

output_zip_LE <- here("data", "life_expectancy.zip")

dir.create("data", showWarnings = FALSE)

GET(url_LE, write_disk(output_zip_LE, overwrite = TRUE))

unzip(output_zip_LE, exdir = here("data"))

file_path_LE<- "data/API_SP.DYN.LE00.IN_DS2_en_csv_v2_87.csv"
life_expectancy_data_raw <- data.table::fread(file_path_LE, encoding = "UTF-8")

new_colnames <- as.character(unlist(life_expectancy_data_raw[1, ]))

setnames(life_expectancy_data_raw, new_colnames)

life_expectancy_data_raw <- life_expectancy_data_raw[-1, ]

setnames(life_expectancy_data_raw, gsub("\\.00000$", "", colnames(life_expectancy_data_raw)))

head(life_expectancy_data_raw)

life_expectancy <- life_expectancy_data_raw[,-c(2,3,4,68,69)]
head(life_expectancy)

tidy_data <- life_expectancy %>%
  pivot_longer(
    cols = starts_with("19") | starts_with("20"), 
    names_to = "Year",                           
    values_to = "LifeExpectancy"                 
  ) %>%
  mutate(Year = as.integer(Year))              

head(tidy_data)
```

###3. Expected Years of Schooling

The dataset for expected years of schooling is sourced from the World Bank's gender data portal. The data is downloaded as a ZIP file and extracted into a local directory. We loaded the extracted CSV file into R and removed unnecessary columns to focus on the relevant data. There were a lot of missing values in the data. So, by using the zoo package, we filled the missing values with the previous non-missing value for consistency. Finally, we filtered the rows where the schooling data is categorized as "total", and renamed the "Value" column to Expected_Schooling for better clarity and consistency among our datasets.

```{r}
# Load necessary libraries
library(httr)  # For downloading files
library(zip)   # For handling ZIP files (optional, built-in utils can also be used)

# Define the URL and output file path
url_ee <- "https://extdataportal.worldbank.org/content/dam/sites/data/gender-data/data/data-gen/zip/indicator/expected-years-of-schooling.zip"
output_zip_ee <- "expected_years_of_schooling.zip"
output_dir_ee <- "expected_years_of_schooling"

# Step 1: Download the file
download.file(url_ee, output_zip_ee, mode = "wb")

# Step 2: Unzip the downloaded file
if (!dir.exists(output_dir_ee)) {
  dir.create(output_dir_ee)
}
unzip(output_zip_ee, exdir = output_dir_ee)

# Step 3: List extracted files
extracted_files <- list.files(output_dir_ee, full.names = TRUE)
print(extracted_files)

# Optional: Load the data (assuming it's in CSV format; adjust as needed)
expected_years_of_schooling_raw<- read.csv("expected_years_of_schooling/Expected years of schooling.csv")
head(expected_years_of_schooling_raw)

expected_years_of_schooling <- expected_years_of_schooling_raw[,-c(1,2,4)]
head(expected_years_of_schooling)
```
```{r}

fill_na_with_previous <- function(x) {
  # Ensure the column is not a factor (convert to character or numeric)
  if (is.factor(x)) x <- as.character(x)
  return(na.locf(x, na.rm = FALSE, fromLast = FALSE))
}

# Apply the function to each column in the dataset
# Use mutate(across()) for all columns or specify columns explicitly
library(zoo) # for na.locf()
expected_years_of_schooling <- expected_years_of_schooling %>%
  mutate(across(everything(), ~ na.locf(., na.rm = FALSE)))

expected_years_of_schooling_total<- expected_years_of_schooling %>% filter(Disaggregation == "total")
expected_years_of_schooling_total[["Year"]] <- as.numeric(as.character(expected_years_of_schooling_total[["Year"]]))
colnames(expected_years_of_schooling_total)[colnames(expected_years_of_schooling_total) == "Value"] <- "Expected_Schooling"
colnames(expected_years_of_schooling_total)[colnames(expected_years_of_schooling_total) == "Country.Name"] <- "Country Name"

head(expected_years_of_schooling_total)


```

###4. GNI per Capita

The Gross National Income (GNI) per capita data is obtained through the World Bank API. We downloaded the data as a CSV file within a ZIP archive. After extraction, we loaded the data into R using fread(). Then, we cleaned and standardized column. We reshaped the dataset into a long format using the melt function, creating a structure where each row represents a country, a year, and its corresponding GNI per capita value to allow for easier integration with other datasets.

```{r}
# Define the World Bank API URL
url_gni <- "https://api.worldbank.org/v2/en/indicator/NY.GNP.PCAP.CD?downloadformat=csv"

# Define the local path to save the file
output_zip_gni <- here("data3", "gni.zip")

# Create the data directory if it doesn't exist
dir.create("data3", showWarnings = FALSE)

# Download the zip file
GET(url_gni, write_disk(output_zip_gni, overwrite = TRUE))

# Unzip the downloaded file
unzip(output_zip_gni, exdir = here("data3"))

# Reload the file with explicit settings
file_path_gni<- "data3/API_NY.GNP.PCAP.CD_DS2_en_csv_v2_4531.csv"
gni_data_raw <- data.table::fread(file_path_gni, encoding = "UTF-8")

# Extract the first row as a character vector for column names
new_colnames <- as.character(unlist(gni_data_raw[1, ]))

# Assign the new column names
setnames(gni_data_raw, new_colnames)

# Remove the first row since it's now the column names
gni_data_raw <- gni_data_raw[-1, ]

# Clean the column names (remove '.00000')
setnames(gni_data_raw, gsub("\\.00000$", "", colnames(gni_data_raw)))

# Check the updated data
head(gni_data_raw)

columns_to_keep <- c(
  "Country Name",
  grep("^197[0-9]|^198[0-9]|^199[0-9]|^200[0-9]|^201[0-9]|^202[0-9]$",
       colnames(gni_data_raw), value = TRUE)
)

gni_data_filtered <- gni_data_raw[, ..columns_to_keep]

# Transform into long format
gni_data_long <- melt(
  gni_data_filtered,
  id.vars = c("Country Name"),
  variable.name = "Year",   # Year column will retain exact year values (e.g., 1970, 1971, etc.)
  value.name = "GNI_Per_Capita"
)

# Ensure "Year" remains a string and convert it to numeric for further processing
gni_data_long[, Year := as.numeric(as.character(Year))]

# Check the updated data
head(gni_data_long)

data.table(gni_data_long)
```

###5. Mean Years of Schooling

The mean years of schooling dataset is downloaded as a CSV file from Our World in Data. After loading the file into R, we again removed the unnecessary columns to retain only the country name, year, and schooling data.

```{r}
# Load necessary library
library(readxl)

# Define the URL
url_ME <- "https://ourworldindata.org/grapher/mean-years-of-schooling-long-run.csv?v=1&csvType=full&useColumnShortNames=true"

# Define the file path to save the CSV locally
file_path_ME <- "mean_years_of_schooling.csv"

# Download the file
download.file(url_ME, file_path_ME)

# Read the CSV file into R
mean_schooling_raw <- read.csv(file_path_ME)
colnames(mean_schooling_raw) <- c("Country Name", "Code", "Year", "MeanYearsOfSchooling")

# Display the first few rows of the dataset
mean_schooling <- mean_schooling_raw[,-2]

head(mean_schooling)
```

###6. Democracy

The political regime dataset, indicating the type of governance in countries over time, is acquired also from Our World in Data. The file is downloaded as a CSV and loaded into R. We removed unnecessary columns to only leave the country name, year, and political regime.

```{r}
# Define the URL
url_D <- "https://ourworldindata.org/grapher/political-regime.csv?v=1&csvType=full&useColumnShortNames=false"

# Define the file path to save the CSV locally
file_path_D <- "democracy.csv"

# Download the file
download.file(url_D, file_path_D)

# Read the CSV file into R
democracy_raw <- read.csv(file_path_D)
colnames(democracy_raw) <- c("Country Name", "Code", "Year", "Political.regime")

# Display the first few rows of the dataset
democracy <- democracy_raw[,-2]
head(democracy)
```

###7. Gender Inequality

The gender inequality index dataset is sourced from the UN Human Development Report through Our World in Data. After downloading the CSV file, the data is loaded into R. Similar to other datasets, irrelevant columns are removed, leaving the country name, year, and gender inequality values.

```{r}
# Define the URL
url_GI <- "https://ourworldindata.org/grapher/gender-inequality-index-from-the-human-development-report.csv?v=1&csvType=full&useColumnShortNames=false"

# Define the file path to save the CSV locally
file_path_GI <- "democracy.csv"

# Download the file
download.file(url_GI, file_path_GI)

# Read the CSV file into R
gender_inequality_raw <- read.csv(file_path_GI)
colnames(gender_inequality_raw) <- c("Country Name", "Code", "Year", "Gender.Inequality")

# Display the first few rows of the dataset
gender_inequality <- gender_inequality_raw[,-2]
head(gender_inequality)
```

###8. Population

The population dataset, reflecting the population size of countries over time, is acquired from Our World in Data. The CSV file is downloaded, loaded into R, and cleaned to exclude unnecessary columns.

```{r}
url_pop <- "https://ourworldindata.org/grapher/population.csv?v=1&csvType=full&useColumnShortNames=true"

file_path_pop <- "population.csv"

download.file(url_pop, file_path_pop)

population_raw <- read.csv(file_path_pop)
colnames(population_raw) <- c("Country Name", "Code", "Year", "Population")

population <- population_raw[,-2] %>% filter(Year > 1960)
head(population)
```

###9. Foreign Direct Investment

The foreign direct investment (FDI) dataset, showing net inflows as a percentage of GDP, is obtained from Our World in Data. The CSV file is downloaded and loaded into R. Again, we dropped irrelevant columns and kept only the essential country name, year, and FDI values. This dataset highlights the economic integration of countries through international investments.

```{r}
url_fdi <- "https://ourworldindata.org/grapher/foreign-direct-investment-net-inflows-as-share-of-gdp.csv?v=1&csvType=full&useColumnShortNames=false"

file_path_fdi <- "FDI.csv"

download.file(url_fdi, file_path_fdi)

fdi_raw <- read.csv(file_path_fdi)
colnames(fdi_raw) <- c("Country Name", "Code", "Year", "FDI.inflow")

fdi <- fdi_raw[,-2]
head(fdi)
```

###10. Tax Revenues 

```{r}
url_tax <- "https://ourworldindata.org/grapher/tax-revenues-as-a-share-of-gdp-unu-wider.csv?v=1&csvType=full&useColumnShortNames=false"

file_path_tax <- "tax_revenues.csv"

download.file(url_tax, file_path_tax)

tax_revenues_raw <- read.csv(file_path_tax)
colnames(tax_revenues_raw) <- c("Country Name", "Code", "Year", "Tax_Share")

tax_revenues <- tax_revenues_raw[,-2]
head(tax_revenues)
```

###11. Electricity Generation
```{r}
url_EG <- "https://ourworldindata.org/grapher/per-capita-electricity-generation.csv?v=1&csvType=full&useColumnShortNames=true"

file_path_EG <- "electricity_generation.csv"

download.file(url_EG, file_path_EG)

electricity_generation_raw <- read.csv(file_path_EG)
colnames(electricity_generation_raw) <- c("Country Name", "Code", "Year", "Electricity_Generation")

electricity_generation <- electricity_generation_raw[,-2]
head(electricity_generation)
```

###12. GDP per Capita
```{r}
url_GDP <- "https://ourworldindata.org/grapher/gdp-per-capita-worldbank.csv?v=1&csvType=full&useColumnShortNames=true"

file_path_GDP <- "electricity_generation.csv"

download.file(url_GDP, file_path_GDP)

GDPpercapita_raw <- read.csv(file_path_GDP)
colnames(GDPpercapita_raw) <- c("Country Name", "Code", "Year", "GDP_per_capita")

GDPpercapita <- GDPpercapita_raw[,-2]
head(GDPpercapita)
```

*Merging All Data

```{r}
# Ensure all datasets are data.tables
tables <- list(
  electricity_generation,
  tax_revenues,
  fdi,
  population,
  gender_inequality,
  democracy,
  mean_schooling,
  gni_data_long,
  expected_years_of_schooling_total,
  tidy_data,
  hdi_long,
  GDPpercapita
)

merged_data <- Reduce(function(x, y) merge(x, y, by = c("Country Name", "Year"), all = TRUE), tables)

merged_data <- subset(merged_data, Year >= 1970 & Year <= 2023)

str(merged_data)

merged_data <- merged_data[!merged_data$`Country Name` %in% c(
"Africa (UN)",                                         
"Africa Eastern and Southern" ,                        
"Africa Western and Central"   ,                       
"Americas (UN)"                 ,                      
"Arab States"                    ,                     
"Arab States (UNDP)"              ,                    
"Arab World"                       ,                   
"Asia"                              ,                  
"Asia (excl. China and India)"       ,                 
"Asia (UN)"                           ,                
"Caribbean small states"               ,               
"Central Europe and the Baltics"        ,              
 "Early-demographic dividend"            ,              
"East Asia & Pacific"                     ,            
 "East Asia & Pacific (excluding high income)"  ,       
  "East Asia & Pacific (IDA & IBRD countries)"   ,       
 "East Asia & Pacific (IDA & IBRD)"               ,     
 "East Asia and Pacific (WB)"                      ,    
 "East Asia and the Pacific"                        ,   
 "East Asia and the Pacific (UNDP)"                  ,  
 "East Germany"                                       , 
 "East Timor"                                          ,
 "Euro area"                                           ,
 "Europe"                                              ,
"Europe (excl. Russia)"                               ,
 "Europe (UN)"                                         ,
 "Europe & Central Asia"                               ,
 "Europe & Central Asia (excluding high income)"       ,
 "Europe & Central Asia (IDA & IBRD countries)"        ,
 "Europe & Central Asia (IDA & IBRD)"                  ,
 "Europe and Central Asia"                             ,
 "Europe and Central Asia (UNDP)"                      ,
 "Europe and Central Asia (WB)"                        ,
 "European Union"                                      ,
 "European Union (27)"                                 ,
 "Fragile and conflict affected situations"            ,
 "Heavily indebted poor countries (HIPC)"              ,
 "High human development"                              ,
 "High human development (UNDP)"                       ,
 "High income"                                         ,
 "High-income countries"                               ,
"IBRD only"                                           ,
 "IDA & IBRD total"                                    ,
 "IDA blend"                                           ,
 "IDA only"                                            ,
 "IDA total"                                           ,
"Late-demographic dividend"                           ,
 "Latin America & Caribbean"                           ,
 "Latin America & Caribbean (excluding high income)"   ,
 "Latin America & Caribbean (IDA & IBRD)"              ,
"Latin America & the Caribbean (IDA & IBRD countries)",
"Latin America and Caribbean (WB)"                    ,
 "Latin America and the Caribbean"                     ,
 "Latin America and the Caribbean (UN)"                ,
 "Latin America and the Caribbean (UNDP)"              ,
"Least developed countries: UN classification"        ,
"Low & middle income"                                 ,
"Low human development"                               ,
"Low human development (UNDP)"                        ,
"Low income"                                          ,
"Low-income countries"                                ,
"Lower middle income"                                 ,
"Lower-middle-income countries"                       ,
"Medium human development"                            ,
"Medium human development (UNDP)"                     ,
"Middle East & North Africa"                          ,
"Middle East & North Africa (excluding high income)"  ,
"Middle East & North Africa (IDA & IBRD countries)"   ,
"Middle East & North Africa (IDA & IBRD)"             ,
"Middle East and North Africa (WB)"                  , 
"Middle income"                                       ,
"Middle-income countries"                             ,
"North America"                                       ,
"North America (WB)"                                  ,
"Northern America (UN)"                               ,
"Not classified"                                      ,
"OECD members"                                        ,
"Other small states"                                  ,
"Pacific island small states"                         ,
"Post-demographic dividend"                           ,
"Pre-demographic dividend"                            ,
"Reunion"                                             ,
"Small states"                                        ,
"South America"                                       ,
"South America (excl. Brazil)"                        ,
"South Asia"                                          ,
"South Asia (IDA & IBRD)"                             ,
"South Asia (UNDP)"                                   ,
"South Asia (WB)"                                     ,
"Sub-Saharan Africa"                                  ,
"Sub-Saharan Africa (excluding high income)"          ,
"Sub-Saharan Africa (IDA & IBRD countries)"           ,
"Sub-Saharan Africa (IDA & IBRD)"                     ,
"Sub-Saharan Africa (UNDP)"                           ,
"Sub-Saharan Africa (WB)"                             ,
"Very high human development"                         ,
"Very high human development (UNDP)"                  ,
"West Bank and Gaza"                                  ,
"Western Sahara"                                      ,
"Oceania" ,                            
"Oceania (UN)"  , 
"World",
"Upper middle income",
"Upper-middle-income countries"
), ]


country_name_mapping <- c(
  "Congo (Democratic Republic of the)" = "Congo, Dem. Rep.",
  "Congo, Rep." = "Congo, Dem. Rep.",
   "Democratic Republic of Congo" =  "Congo, Rep." ,
   "Democratic Republic of Congo" = "Congo",
  "Congo, Dem. Rep."="Congo",
  "Bahamas, The"="Bahamas",
    "C\xf4te d'Ivoire"="Cote d'Ivoire",
  "Czechia"="Czechoslovakia",
  "Gambia, The"= "Gambia" ,
  "Bolivia (Plurinational State of)"= "Bolivia",
  "Ethiopia (former)"="Ethiopia",
  "Egypt, Arab Rep."="Egypt",
  "Hong Kong"="Hong Kong SAR, China",
  "Hong Kong SAR, China" ="Hong Kong, China (SAR)",
  "Eswatini (Kingdom of)"="Eswatini", 
   "Iran (Islamic Republic of)"="Iran",
 "Iran, Islamic Rep."="Iran",
  "Korea (Democratic People's Rep. of)" = "Korea (Republic of)",
  "Korea (Republic of)"="Korea, Dem. People's Rep.",
  "Korea, Dem. People's Rep."=    "Korea, Rep.",
   "Micronesia (country)"="Micronesia (Federated States of)" ,
   "Micronesia (Federated States of)"="Micronesia, Fed. Sts.",
  "Micronesia, Fed. Sts."="Micronesia",
"Moldova (Republic of)"="Moldova" ,
"Kyrgyz Republic"= "Kyrgyzstan"    ,
"Lao PDR"="Lao People's Democratic Republic"   ,
 "Macao"="Macao SAR, China",
"Palestine"="Palestine, State of",
   "Palestine, State of"="Palestine/Gaza",
   "Palestine/Gaza" = "Palestine/West Bank",
"Palestine/West Bank"  = "Palestine/Gaza" ,
"Slovak Republic"="Slovakia"                    ,
"Yemen" ="Yemen Arab Republic",
   "Yemen Arab Republic"="Yemen People's Republic",
   "Yemen People's Republic"="Yemen, Rep.",
"Russia"="Russian Federation",
"Serbia"="Serbia and Montenegro"          ,
"Syrian Arab Republic" ="Syria"         ,
  "Viet Nam"= "Vietnam",     
  "Democratic Republic of Vietnam" = "Republic of Vietnam",
 "Tanzania (United Republic of)"="Tanzania" ,
"T\xfcrkiye" = "Turkey",
"Turkey"="Turkiye",
 "Venezuela (Bolivarian Republic of)" = "Venezuela",
"Venezuela, RB" = "Venezuela"

)
new_data <- merged_data %>%
  mutate(`Country Name` = recode(`Country Name`, !!!country_name_mapping))

country_name_mapping2 <- c(
  "Palestine/West Bank"  = "Palestine/Gaza",
  "Congo, Dem. Rep."    = "Congo",
  "Turkey" = "Turkiye",
   "Yemen, Rep." = "Yemen People's Republic",
  "Micronesia, Fed. Sts."    =  "Micronesia" ,
   "Korea, Rep."     = "Korea, Dem. People's Rep."  
)
  
new_data <- new_data %>%
  mutate(`Country Name` = recode(`Country Name`, !!!country_name_mapping2))


str(new_data)

# Group by Country Name and Year, and summarize other columns
new_data <- new_data %>%
  group_by(`Country Name`, Year) %>%
  summarize(across(everything(), ~ mean(., na.rm = TRUE)), .groups = "drop")

str(new_data)


```

```{r}
hdicode_group <- hdi_long[, .(
  Number_of_Countries = uniqueN(`Country Name`) 
), by = hdicode]

# View the grouped dataset
print(hdicode_group)

set.seed(213276)
low <- hdi_long[hdicode == "Low", sample(unique(`Country Name`), 5)]

set.seed(213276)
medium <- hdi_long[hdicode == "Medium", sample(unique(`Country Name`), 5)]

set.seed(213276)
high <- hdi_long[hdicode == "High", sample(unique(`Country Name`), 5)]

set.seed(213276)
very_high <- hdi_long[hdicode == "Very High", sample(unique(`Country Name`), 5)]

# Subset new_data to include only the selected countries
selected_countries <- new_data %>%
  filter(`Country Name` %in% c(low, medium, high, very_high))

# Add an HDI category column for plotting
selected_countries <- selected_countries %>%
  mutate(
    HDI_Group = case_when(
      `Country Name` %in% low ~ "Low",
      `Country Name` %in% medium ~ "Medium",
      `Country Name` %in% high ~ "High",
      `Country Name` %in% very_high ~ "Very High"
    )
  )
```
Life Expectancy Plot
```{r}
ggplot(new_data, aes(x = Year, y = LifeExpectancy, color = LifeExpectancy, group = `Country Name`)) +
  geom_line(alpha = 0.7) +
  scale_color_viridis_c() +
  theme_minimal()
```


```{r}
library(ggplot2)
library(ggrepel)

label_data <- selected_countries %>%
  group_by(`Country Name`, HDI_Group) %>%
  filter(Year == 2022)

# Plot
ggplot(selected_countries, aes(x = Year, y = LifeExpectancy, color = `Country Name`, group = `Country Name`)) +
  geom_line(size = 0.5) +
  geom_text_repel(
    data = label_data,
    aes(label = `Country Name`),
    show.legend = FALSE,
    size = 2
  ) +
  facet_wrap(~ HDI_Group, ncol = 2, scales = "fixed") +  # Separate facets for each group
  labs(
    title = "Life Expectancy Over Time by HDI Group",
    subtitle = "Selected Countries",
    x = "Year",
    y = "Life Expectancy"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "gray30"),
    legend.position = "none",  # Remove legend
    panel.grid.minor = element_blank()
  )

```


```{r}
# Ensure data is ordered by Country and Year
new_data <- new_data %>% arrange(`Country Name`, Year)

# Function to create interactive plot for a group of countries
create_interactive_plot <- function(data, group, title) {
  filtered_data <- data %>% filter(`Country Name` %in% group)

  plot <- ggplot(filtered_data, aes(
    x = Year, y = LifeExpectancy,
    color = `Country Name`,  # Use country names for distinct colors
    group = `Country Name`,  # Ensure proper grouping for lines
    text = paste("Country: ", `Country Name`, "<br>",
                 "Year: ", Year, "<br>",
                 "Life Expectancy: ", round(LifeExpectancy, 2))
  )) +
    geom_line(size = 0.5, alpha = 0.8) +  # Ensure proper line plotting
    theme_minimal() +
    scale_color_viridis_d() +  # Use discrete viridis colors for countries
    labs(
      title = title,
      x = "Year",
      y = "Life Expectancy (Years)"
    )

  ggplotly(plot, tooltip = "text")
}

# Create and print interactive plots for each group
interactive_low <- create_interactive_plot(
  new_data, low, "Life Expectancy Over Time in Countries with Low HDI Scores"
)
interactive_low

interactive_high <- create_interactive_plot(
  new_data, high, "Life Expectancy Over Time in Countries with High HDI Scores"
)
interactive_high

interactive_medium <- create_interactive_plot(
  new_data, medium, "Life Expectancy Over Time in Countries with Medium HDI Scores"
)
interactive_medium

interactive_very_high<- create_interactive_plot(
  new_data, very_high, "Life Expectancy Over Time in Countries with Very High HDI Scores"
)
interactive_very_high
```


Expected Years of Schooling Plot
```{r}
ggplot(new_data, aes(x = Year, y = Expected_Schooling, color = Expected_Schooling, group = `Country Name`)) +
  geom_line(alpha = 0.7) +
  scale_color_viridis_c() +
  theme_minimal()
```

```{r}
ggplot(selected_countries, aes(x = Year, y = Expected_Schooling, color = `Country Name`, group = `Country Name`)) +
  geom_line(size = 0.5) +
  geom_text_repel(data = label_data, aes(label = `Country Name`), hjust = -0.2, show.legend = TRUE) +
  facet_wrap(~ HDI_Group, ncol = 2, scales = "fixed") +  # 2x2 layout
  labs(
    title = "Expected Years of Schooling Over Time by HDI Group",
    subtitle = "Selected Countries",
    x = "Year",
    y = "Expected Years of Schooling",
    color = "Country"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "gray30"),
    legend.position = "none",  # Remove legend as labels are on the plot
    panel.grid.minor = element_blank()
  )
```

GNI plots
```{r}
library(ggplot2)
library(RColorBrewer)
library(dplyr)
library(ggrepel)

# Define color palettes for each group
colors_very_high <- brewer.pal(5, "Reds")      # Reds for very high
colors_high <- brewer.pal(5, "Blues")          # Blues for high
colors_medium <- brewer.pal(5, "Greens")       # Greens for medium
colors_low <- brewer.pal(5, "Purples")         # Purples for low

# Assign colors to countries based on their group
country_colors <- c(
  setNames(colors_low, low),                   # Map Low group to Purples
  setNames(colors_medium, medium),             # Map Medium group to Greens
  setNames(colors_high, high),                 # Map High group to Blues
  setNames(colors_very_high, very_high)        # Map Very High group to Reds
)

# Add Group column to gni_data_long based on existing vectors
gni_data_long <- gni_data_long %>%
  mutate(
    Group = case_when(
      `Country Name` %in% low ~ "Low",
      `Country Name` %in% medium ~ "Medium",
      `Country Name` %in% high ~ "High",
      `Country Name` %in% very_high ~ "Very High",
      TRUE ~ NA_character_  # Explicitly set NA for unmatched countries
    )
  ) %>%
  filter(!is.na(Group))  # Remove rows with NA in Group

# Find the last year for each country to position labels
label_data <- gni_data_long %>%
  group_by(`Country Name`, Group) %>%
  filter(Year == max(Year))

# Line plot comparing all groups with specific colors
ggplot(selected_countries, aes(x = Year, y = GNI_Per_Capita, color = `Country Name`, group = `Country Name`)) +
  geom_line(size = 0.8) +
  geom_text_repel(data = label_data, aes(label = `Country Name`), hjust = -0.2, show.legend = FALSE) +
  facet_wrap(~ Group, scales = "free_y") +  # Separate facets for each group
  scale_color_manual(values = country_colors) +  # Apply custom colors
  theme_minimal() +
  labs(
    title = "GNI Per Capita Over Time by Group",
    x = "Year",
    y = "GNI Per Capita (USD)"
  ) +
  theme(
    legend.position = "none",  # Remove legend
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "gray30")
  )

```

Mean Schooling Plot
```{r}
ggplot(mean_schooling, aes(x = Year, y = MeanYearsOfSchooling, color = MeanYearsOfSchooling, group = `Country Name`)) +
  geom_line(alpha = 0.7) +
  scale_color_viridis_c() +
  theme_minimal()
```

democracy plot

```{r}
# Summarize the data
aggregated_data <- democracy %>%
  group_by(Year, Political.regime) %>%
  summarize(Count = n(), .groups = "drop")

# Create the line graph
ggplot(aggregated_data, aes(x = Year, y = Count, color = factor(Political.regime), group = Political.regime)) +
  geom_line(size = 0.5) +
  geom_point(size = 0.2) +  # Optional: Add points for emphasis
  scale_color_viridis_d() +
  theme_minimal(base_size = 14) +
  labs(
    title = "Number of Countries by Political Regime Over Time",
    x = "Year",
    y = "Number of Countries",
    color = "Political Regime"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for readability
    legend.position = "right"  # Move the legend to the side
  )

```


gender inequality

```{r}
ggplot(gender_inequality, aes(x = Year, y = Gender.Inequality, color = Gender.Inequality, group = `Country Name`)) +
  geom_line(alpha = 0.7) +
  scale_color_viridis_c() +
  theme_minimal()

```

population

```{r}
ggplot(population, aes(x = Year, y = Population, color = Population, group = `Country Name`)) +
  geom_line(alpha = 0.7) +
  scale_color_viridis_c() +
  theme_minimal()
```


fdi
```{r}
ggplot(fdi, aes(x = Year, y = FDI.inflow, color = FDI.inflow, group = `Country Name`)) +
  geom_line(alpha = 0.7) +
  scale_color_viridis_c() +
  theme_minimal()
```


gdp
```{r}
ggplot(new_data, aes(x = Year, y = GDP_per_capita, color = GDP_per_capita, group = `Country Name`)) +
  geom_line(alpha = 0.7) +
  scale_color_viridis_c() +
  theme_minimal()
```

tax
```{r}
ggplot(tax_revenues, aes(x = Year, y = Tax_Share, color = Tax_Share, group = `Country Name`)) +
  geom_line(alpha = 0.7) +
  scale_color_viridis_c() +
  theme_minimal()
```

electricity

```{r}
ggplot(electricity_generation, aes(x = Year, y = Electricity_Generation, color = Electricity_Generation, group = `Country Name`)) +
  geom_line(alpha = 0.7) +
  scale_color_viridis_c() +
  theme_minimal()
```




```{r}

necessary_columns <- c("Country Name", "Year", "LifeExpectancy", "GNI_Per_Capita", 
                       "MeanYearsOfSchooling", "Expected_Schooling", "HDI")

# Select only the gerekli columns from merged_data
model_data <- new_data %>%
  select(all_of(necessary_columns))

model <- lm(HDI  ~ LifeExpectancy + GNI_Per_Capita + MeanYearsOfSchooling + Expected_Schooling, data=model_data )

summary(model)

# Specify the columns to check
columns_to_check <- c("LifeExpectancy", "GNI_Per_Capita", "MeanYearsOfSchooling", "Expected_Schooling", "HDI")

# Identify rows where exactly one column is missing
rows_with_one_missing <- model_data %>%
  rowwise() %>%
  mutate(missing_count = sum(is.na(c_across(all_of(columns_to_check))))) %>%
  filter(missing_count == 1) %>%
  ungroup()

# View the filtered dataset
head(rows_with_one_missing)

# Number of rows with exactly one missing value
nrow(rows_with_one_missing)

# Regression coefficients from your model
coefficients <- coef(model)
```
```{r}
# Define a function to predict the missing value or calculate HDI
predict_or_calculate_hdi <- function(row, coefficients) {
  intercept <- coefficients["(Intercept)"]
  
  # Initialize the prediction with the intercept
  predicted_value <- intercept
  
  # Add contributions from the available independent variables
  if (!is.na(row$LifeExpectancy)) {
    predicted_value <- predicted_value + coefficients["LifeExpectancy"] * row$LifeExpectancy
  }
  if (!is.na(row$GNI_Per_Capita)) {
    predicted_value <- predicted_value + coefficients["GNI_Per_Capita"] * row$GNI_Per_Capita
  }
  if (!is.na(row$MeanYearsOfSchooling)) {
    predicted_value <- predicted_value + coefficients["MeanYearsOfSchooling"] * row$MeanYearsOfSchooling
  }
  if (!is.na(row$Expected_Schooling)) {
    predicted_value <- predicted_value + coefficients["Expected_Schooling"] * row$Expected_Schooling
  }
  
  return(predicted_value)
}

# Apply the function to fill in the missing values and calculate HDI
completed_rows <- rows_with_one_missing %>%
  rowwise() %>%
  mutate(
    LifeExpectancy = ifelse(
      is.na(LifeExpectancy),
      predict_or_calculate_hdi(cur_data(), coefficients),
      LifeExpectancy
    ),
    GNI_Per_Capita = ifelse(
      is.na(GNI_Per_Capita),
      predict_or_calculate_hdi(cur_data(), coefficients),
      GNI_Per_Capita
    ),
    MeanYearsOfSchooling = ifelse(
      is.na(MeanYearsOfSchooling),
      predict_or_calculate_hdi(cur_data(), coefficients),
      MeanYearsOfSchooling
    ),
    Expected_Schooling = ifelse(
      is.na(Expected_Schooling),
      predict_or_calculate_hdi(cur_data(), coefficients),
      Expected_Schooling
    ),
    # Calculate HDI for all rows using the function
    HDI = ifelse(
      is.na(HDI),
      predict_or_calculate_hdi(cur_data(), coefficients),
      HDI
    )
  ) %>%
  ungroup()

# View the updated completed_rows dataset
head(completed_rows)
```

```{r}
library(dplyr)

# Merge predicted values back into model_data
model_data_with_predictions <- model_data %>%
  left_join(
    completed_rows %>%
      select(`Country Name`, Year, LifeExpectancy, GNI_Per_Capita, MeanYearsOfSchooling, Expected_Schooling, HDI),
    by = c("Country Name", "Year"),
    suffix = c("_original", "_predicted")
  ) %>%
  mutate(
    # Replace missing values in the original dataset with predicted values
    LifeExpectancy = ifelse(is.na(LifeExpectancy_original), LifeExpectancy_predicted, LifeExpectancy_original),
    GNI_Per_Capita = ifelse(is.na(GNI_Per_Capita_original), GNI_Per_Capita_predicted, GNI_Per_Capita_original),
    MeanYearsOfSchooling = ifelse(is.na(MeanYearsOfSchooling_original), MeanYearsOfSchooling_predicted, MeanYearsOfSchooling_original),
    Expected_Schooling = ifelse(is.na(Expected_Schooling_original), Expected_Schooling_predicted, Expected_Schooling_original),
    HDI = ifelse(is.na(HDI_original), HDI_predicted, HDI_original)
  ) %>%
  # Drop intermediate columns for clarity
  select(-ends_with("_original"), -ends_with("_predicted"))


# Check the structure to confirm integration
str(model_data_with_predictions)

model2 <- lm(HDI  ~ LifeExpectancy + GNI_Per_Capita + MeanYearsOfSchooling + Expected_Schooling, data=model_data_with_predictions )
summary(model2)
```

```{r}
library(dplyr)
library(ggplot2)

# Calculate the number of cells filled for each column
filled_counts <- rows_with_one_missing %>%
  summarise(
    LifeExpectancy = sum(is.na(LifeExpectancy) & !is.na(completed_rows$LifeExpectancy)),
    GNI_Per_Capita = sum(is.na(GNI_Per_Capita) & !is.na(completed_rows$GNI_Per_Capita)),
    MeanYearsOfSchooling = sum(is.na(MeanYearsOfSchooling) & !is.na(completed_rows$MeanYearsOfSchooling)),
    Expected_Schooling = sum(is.na(Expected_Schooling) & !is.na(completed_rows$Expected_Schooling)),
    HDI = sum(is.na(HDI) & !is.na(completed_rows$HDI))
  ) %>%
  pivot_longer(cols = everything(), names_to = "Column", values_to = "FilledCount")

# Create a bar plot
ggplot(filled_counts, aes(x = Column, y = FilledCount, fill = Column)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Number of Cells Filled for Each Column",
    x = "Column",
    y = "Number of Cells Filled",
    fill = "Column"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```

# Decision Trees
###  Separating GDP into 5 Buckets
At first we tried to predict GNI, however GNI data was very limited. Therefore, we chaged our approach and decided to use GDP per capita as a mesure of wealth of the country. 
We created 4 buckets of GDP per capita. We used percentiles to separate GDP per capita into equel size buckets.We categorized GDP per capita in countries in different years as low, lower middle, upper middle and high.

```{r}
percentiles <- quantile(new_data$GDP_per_capita, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)
new_data$GDP_Bucket <- cut(
  new_data$GDP_per_capita,
  breaks = percentiles,
  labels = c("Low", "Lower-Middle", "Upper-Middle-Low", "High"),
  include.lowest = TRUE
)

table(new_data$GDP_Bucket)
```
#### Baseline Guessing
To be able to compare our decision tree models' performances, we printed the random guessing model's accuracy. 
```{r} 
baseline_random <- 1 / 4
print(paste("Baseline Accuracy (Random Guessing):", baseline_random))
```

## Decision Tree with only HDI Variables
We want to see if the factors in HDI like life expectancy and education has a role on GNI. Therefore, we regressed GDP classes on mean years of schooling, expected years of schooling and life expectancy. 

```{r}
set.seed(375)  
split <- sample(1:nrow(new_data), size = 0.7 * nrow(new_data))
train_data <- new_data[split, ]
test_data <- new_data[-split, ]
```

```{r}
library(rpart.plot)
library(rpart)
tree_model_HDI <- rpart(
  GDP_Bucket ~ MeanYearsOfSchooling + Expected_Schooling + LifeExpectancy,
  data = train_data,
  method = "class", 
  #control = rpart.control(minsplit = 5)
)
visNetwork::visTree(tree_model_HDI)

```
The most prominant factor determining the GDP per capita is life expectancy. However, this only shows us a strong correlation between life expectancy and GDP per capita. We can not determine the direction of the relationship from this decision tree. 

```{r}
predictions_HDI <- predict(tree_model_HDI, test_data, type = "class")
confusion_matrix_HDI <- caret::confusionMatrix(table(fitted = predictions_HDI, actual = test_data$GDP_Bucket))
print(confusion_matrix_HDI)
```
Accuracy is 0.63 which a lot higher than baseline accuracy. Therefore, our model improves the predictions. However, its success is limited. 
We are going to incorparate more variables into our model to improve its prediction power.

## Continuous GDP Prediction
Without separating GDP into buckets, we used a continuous decision tree model to predict the exact value of GDP per capita.
We wanted to see if there are any differences on most effective factors between continuous and discrete GDP per capita values. 

### Decision Tree with All variables minus Electricty Generation

```{r}
library(rpart.plot)
library(rpart)
tree_model_all <- rpart(
  GDP_per_capita ~  Tax_Share + Population +Gender.Inequality + Political.regime + MeanYearsOfSchooling + Expected_Schooling + LifeExpectancy + FDI.inflow,
  data = train_data,
  method = "anova", 
  #control = rpart.control(minsplit = 5)
)

visNetwork::visTree(tree_model_all)
```

#### Decision Tree with All variables after 2000
Electricity generation data before 2000 was limited for most of the countries. Therefore, we wanted to train another decision tree model with the data after 2000 to see if electricity generation is an important factor of GDP per capita.

```{r}
library(dplyr)
library(tidyr)

data_after_2000 <- new_data %>%
  filter(Year >= 2000)

head(data_after_2000)
```
We separated the data after 2000 into training and test sets.

```{r}
set.seed(389)  
split_2000 <- sample(1:nrow(data_after_2000), size = 0.7 * nrow(data_after_2000))
train_data_after2000 <- data_after_2000[split_2000, ]
test_data_after2000 <- data_after_2000[-split_2000, ]
```

```{r}
tree_model_2000 <- rpart(
  GNI_Per_Capita ~  Tax_Share + Population + Electricity_Generation + Gender.Inequality + Political.regime + MeanYearsOfSchooling +
    Expected_Schooling + LifeExpectancy + FDI.inflow,
  data = train_data_after2000,
  method = "anova", 
  #control = rpart.control(minsplit = 5)
)

visNetwork::visTree(tree_model_2000)
```

Using the data after 2000 showed us that the most important factor to determine GDP per capita is population. After that electricity generation is the factor that decreases the entropy the most. Higher electricity generation indicated a higher GDP per capita. This is perfectly logical since most of the advance economies use high amounts of electricity in production and technology services. 

## Decision Tree with 4 GDP Buckets
Now instead of using a continuous model, we are going to use a discrete classification model with 4 groups of the target variable. 

### With All the Data

```{r}
library(rpart)
library(rpart.plot)

tree_model_buckets <- rpart(
  GDP_Bucket ~ Electricity_Generation + Tax_Share + Population + Gender.Inequality + Political.regime + MeanYearsOfSchooling + Expected_Schooling + LifeExpectancy ,
  data = train_data,
  method = "class", 
  control = rpart.control(minsplit = 20, cp = 0.01) )

rpart.plot(tree_model_buckets, type = 3, extra = 104, box.palette = "Blues", shadow.col = "gray")

```

```{r}
visNetwork::visTree(tree_model_buckets)
```

In the discrete model, most important factor is life expectancy. We now that most developed nations have higher life expectancies therefore higher life expectancy points to a higher GDP. Again we see the importance of electricity generation at it is the second and third node.
The significant difference between discrete and continuous model is that in the discrete one, population is not that important to determine GDP as it is the factor in the last node.

```{r}
predictions_buckets_all <- predict(tree_model_buckets, test_data, type = "class")
confusion_matrix_buckets_all <- caret::confusionMatrix(table(fitted = predictions_buckets_all, actual = test_data$GDP_Bucket))
print(confusion_matrix_buckets_all)
```
We see that accuracy increased to 73%. It increased 10 percentage point compared to using only HDI factors. Incorparating more factors into the model allows us to make better GDP per capita predictions.

### After 2000 with buckets of GDP 
We are going to do a decision tree with all data after 2000 to predict the group of GDP per capita.
```{r}
tree_model_buckets_2000 <- rpart(
  GDP_Bucket ~ Electricity_Generation + Tax_Share + Population + Gender.Inequality + Political.regime + MeanYearsOfSchooling + Expected_Schooling + LifeExpectancy ,
  data = train_data_after2000,
  method = "class", 
  control = rpart.control(minsplit = 20, cp = 0.01)  # 
)

visNetwork::visTree(tree_model_buckets_2000)
```
Again, countries producing high electricity are classified as high GDP per capita countries. And high life expectancy points to a high GDP. 
The effect of political regime is an interesting one. As Daron Acemoğlu mentioned in his books, democracy is an important factor in development and GDP. He said that countries with storng democracies are generally richer. However in our model the effet of political regime is the opposite one. Countries with 0 in their political regimes are autocracies. And in this model, 0 in political regime points to a high GDP.  

```{r}
predictions_buckets_2000 <- predict(tree_model_buckets, test_data_after2000, type = "class")
confusion_matrix_buckets_2000 <- caret::confusionMatrix(table(fitted = predictions_buckets_2000, actual = test_data_after2000$GDP_Bucket))

print(confusion_matrix_buckets_2000)
```

The accuracy of the model that is trained with the data after 2000 is higher with 75% score. This may be due to the fact that the data quality is higher and there are much less unavaliable data points after 2000. Therefore, the model can make better predictions and has a lot more data to learn from. 
